The Construction of Set-Truncated Higher Inductive Types

### 2021-11-17 paper/paper.tex

% Endpoints
\newcommand{\ep}[3]{\mathcal{E}_{#1}(#2,#3)} % type of endpoinst
\newcommand{\id}[1]{\constructor{id}_{#1}} % identity endpoints
\newcommand{\comp}[2]{#1 \cdot #2} % composition of endpoints
\newcommand{\inle}{\constructor{inl}} %left inclusion of endpoints
\newcommand{\inre}{\constructor{inr}} % right inclusion of endpoints
\newcommand{\prle}{\constructor{pr}_1} % first projection of endpoints
\newcommand{\prre}{\constructor{pr}_2} % second projection of endpoints
\newcommand{\pair}[2]{(#1 , #2)} % pairing of endpoints
\newcommand{\Ce}{\constructor{c}} % constant endpoint
\newcommand{\constr}{\constructor{constr}} % constructor endpoint

@@

% Algebras
\newcommand{\semP}[1]{\llbracket #1 \rrbracket} % semantics of polynomials
\newcommand{\prealg}[1]{\category{PreAlg}_{\hset}(#1)} % category of prealgebras
\newcommand{\forget}[1]{\mathcal{F}_{#1}} % forgetful functor
\newcommand{\semE}[1]{\llbracket #1 \rrbracket} % semantics of endpoints

\newcommand{\Alg}[1]{\category{Alg}_{\hset}(#1)} % category of algebras
\newcommand{\constrA}[1]{\function{map}_{#1}} % point constructor of algebra
\newcommand{\constrP}[1]{\function{path}_{#1}} % path constructor of algebra

@@

% Examples of HITs
\newcommand{\Circ}{S^1} % circle
\newcommand{\base}{\constructor{base}} % base point of circle
\newcommand{\loopS}{\constructor{loop}} % base point of circle
\newcommand{\modZ}{\mathbb{Z}_2} % integers modulo 2
\newcommand{\ZZ}{\constructor{Z}} % zero
\newcommand{\SZ}{\constructor{S}} % successor
\newcommand{\modP}{\constructor{mod}} % n = n + 2
\newcommand{\HIT}{\function{HIT}} % the HIT of a signature

@@

\def\lastname{Van der Weide, Geuvers}
\begin{document}
\begin{frontmatter}
  \title{The Construction of Set-Truncated Higher Inductive Types} \author{Niels van der Weide
		\thanksref{ALL}\thanksref{myemail}}
  \address{Institute for Computation and Information Sciences\\ Radboud Universiteit\\
  	Nijmegen, The Netherlands}
  \author{Herman Geuvers\thanksref{coemail}}
  \address{Institute for Computation and Information Sciences\\ Radboud Universiteit\\
  	Nijmegen, The Netherlands}
  \thanks[ALL]{The author would like to thank Dan Frumin and Andrej Bauer for inspiring discussions. Work on this article was supported by a grant from the \href{https://eutypes.cs.ru.nl/}{COST Action EUTypes CA15123}.} \thanks[myemail]{Email:
  	\href{mailto:nweide@cs.ru.nl} {\texttt{\normalshape
  			nweide@cs.ru.nl}}} \thanks[coemail]{Email:
  	\href{mailto:herman@cs.ru.nl} {\texttt{\normalshape
  			herman@cs.ru.nl}}}
\begin{abstract} 
  We construct finitary set-truncated higher inductive types (HITs) from quotients and the propositional truncation.
  For that, we first define signatures as a modification of the schema by Basold \etal, and we show they give rise to univalent categories of algebras in both sets and setoids.
  To interpret HITs, we use the well-known method of initial algebra semantics.
  The desired algebra is obtained by lifting the quotient adjunction to the level of algebras and adapting Dybjer's and Moeneclaey's interpretation of HITs in setoids.
  From this construction, we conclude that the equality types of HITs are freely generated and that HITs are unique.
  The results are formalized in the UniMath library.
\end{abstract}
\begin{keyword}
higher inductive types, homotopy type theory, category theory, setoids, intuitionistic type theory, Coq
\end{keyword}
\end{frontmatter}
\section{Introduction}
\label{intro}
Homotopy type theory (HoTT) is a form of intensional type theory. 
It has semantics in the simplicial sets model \cite{simpset}, and types represent spaces, terms represent points, and equalities represent paths, and so on.
Furthermore, equality is proof relevant and we frequently talk about homotopies: paths between paths.

One of the main features of HoTT, is higher inductive types (HITs) \cite{hottbook}.
These are types generated by constructors for their points, paths, homotopies, and so on.
HITs have been used in numerous applications among which are homotopical patch theory \cite{AngiuliMLH16}. synthetic homotopy theory \cite{LicataF14,LicataS13,hottbook}, defining type theory within type theory \cite{AltenkirchK16}, constructive finiteness \cite{FruminGGW18}, and the partiality monad \cite{AltenkirchDK17}.
To get a feeling for what HITs are, let us look at some examples.

@@

\begin{lstlisting}[mathescape=true]
Inductive $\Circ$ :=
| $\base$ : $\Circ$
| $\loopS$ : $\base = \base$
\end{lstlisting}
&
\begin{lstlisting}[mathescape=true]
Inductive $\trunct{A}$ :=
| $\truncel{\cdot}$ : $A \rightarrow \trunct{A}$
| $p$ : $\deprod{x, y}{\trunct{A}}{x = y}$
\end{lstlisting}
&
\begin{lstlisting}[mathescape=true]
Inductive $\modZ$ :=
| $\ZZ$ : $\modZ$
| $\SZ$ : $\modZ \rightarrow \modZ$
| $\modP$ : $\deprod{x}{\modZ}{\SZ(\SZ \> x) = x}$
\end{lstlisting}
\end{tabular}

The first one, $\Circ$, is the \emph{circle}.
It has a point $\base$ and a path $\loopS : \base = \base$.
Since neither the point nor path constructors uses arguments from $\Circ$, this HIT is \emph{nonrecursive}.
The second, $\trunct{A}$, is the \emph{propositional truncation} of $A$.
This type is $A$ with all its points identified.
Note that the path constructor $p$ uses arguments from $\trunct{A}$, which means this HIT is \emph{recursive}.
The last one, $\modZ$, is the \emph{integers modulo 2} and note that both the point constructor $\SZ$ and path constructor $\modP$ are recursive.
A HIT is called \emph{finitary} if its point constructor is described by a finitary polynomial and it is called \emph{set-truncated} it is a set.

@@

A similar result was obtained by Kaposi \etal \ \cite{KaposiKA19}, but we do not assume UIP and we use simpler HITs for the construction.
Furthermore, we use this construction to deduce several properties of HITs such as uniqueness and that the path space is freely generated.
Kraus and Von Raumer obtained a similar property for the coequalizer HIT by providing a nicer induction principle for its path types \cite{KrausRaumer}.

\subsection{Background and Related Work}
HITs have been studied formally and several schemes have already been developed.
Awodey \etal \ showed that the rules of inductive types \cite{dybjer1994inductive} in HoTT are equivalent to being a homotopy initial algebra \cite{AwodeyGS12}.
They extended this result to a class of HITs called \emph{W-suspensions} \cite{Sojakova15}.
Another scheme, quotient inductive-inductive types (QIITs), was defined semantically by Altenkirch \etal \ and they show initial algebras are QIITs \cite{AltenkirchCDKF18}.
Assuming UIP, Kaposi \etal \ constructed finitary QIITs as the syntax of a domain specific type theory \cite{KaposiKA19}, and they extended this scheme to higher inductive-inductive types \cite{KaposiK18}.
Basold \etal \ defined an external syntactic scheme for HITs with only point and path constructors \cite{BasoldGW17}
while Dybjer and Moeneclaey defined a scheme allowing constructors for points, paths, and homotopies, and they give its interpretation in the groupoid model  \cite{DybjerM18,HofmannS94}.
Cavallo and Harper extend computational cubical type theory with a scheme for indexed cubical inductive types and proved canonicity \cite{AngiuliHW17,CavalloH19}.

Furthermore, the semantics of HITs also has been studied.
Coquand \etal \ provide the semantics of various higher inductive types, such as spheres, the torus, the truncation, and the pushout, in cubical type theory \cite{BezemCH13,CohenCHM16,CoquandHM18}.
On the other hand, Lumsdaine and Shulman also study the semantics of higher inductive types using a semantic scheme (cell monads with parameters) and they prove existence in sufficiently nice Quillen model categories \cite{lumsdaine2017semantics}.
Some recursive higher inductive types can be constructed from nonrecursive HITs.
Both Kraus and Van Doorn show that the propositional truncation can be constructed from non-recursive HITs \cite{Kraus16,Doorn16} and Rijke shows that every truncation can be constructed via non-recursive HITs \cite{rijke2017join}.

@@

All material in this paper are formalized over the UniMath library \cite{UniMath}.
The formalization is 4166 lines of code and, together with additional examples, it can be found on \url{https://github.com/nmvdw/SetHITs}.

@@

\begin{definition}
\label{def:setoid}
A \emph{setoid} is a set $X$ together with an equivalence relation on $X$.
A \emph{setoid morphism} between two setoids  is a map between the underlying sets which preserves the equivalence relation.
The \emph{category of setoids} $\setoids$ is the category with setoids and setoid morphisms as objects and morphisms.
\end{definition}

@@

Note that in the usual category of setoids, one would take a quotient of the setoid morphisms.
However, in our construction, we do not need any additional equality and so, we just use the ordinary one.

If $X$ is a set and $R$ is an equivalence relation on $X$, then we write $(X, R)$ for the setoid with this data.
If no confusion arises, we write $x \equiv y$ for $R \> x \> y$.
We finish this section by giving some operations of functors and natural transformations, which we need later.
Functors and natural transformations are defined the usual way \cite{mac2013categories}, and we write $\functor{\mathcal{C}}{\mathcal{D}}$ and $\nattrans{F}{G}$ for the type of functors from $\mathcal{C}$ to $\mathcal{D}$ and transformations from $F$ to $G$ respectively.

\begin{notation}
\label{def:functor}
We have the following functors.
\begin{itemize}
	\item For each category $\mathcal{C}$, we have the identity $\idf{\mathcal{C}} : \functor{\mathcal{C}}{\mathcal{C}}$.
	\item For $F : \functor{\mathcal{C}_1}{\mathcal{C}_2}$ and $G : \functor{\mathcal{C}_2}{\mathcal{C}_3}$, we have a composition $\compf{F}{G} : \functor{\mathcal{C}_1}{\mathcal{C}_2}$;
	\item If $\mathcal{D}$ has binary products, then for $F, G : \functor{\mathcal{C}}{\mathcal{D}}$, we have a product $\prodf{F}{G} : \functor{\mathcal{C}}{\mathcal{D}}$;
	\item If $\mathcal{D}$ has binary sums, then for $F, G : \functor{\mathcal{C}}{\mathcal{D}}$, we have a sum $\sumf{F}{G} : \functor{\mathcal{C}}{\mathcal{D}}$;
	\item For each object $X : \mathcal{D}$, we have a constant functor $\Cf{X} : \functor{\mathcal{C}}{\mathcal{D}}$.
\end{itemize}
\end{notation}

@@

\section{Signature of HITs}
\label{sec:signature}
Before we study the construction of set truncated HITs, we first give a precise definition of those.
We do this by internalizing the scheme by Basold \etal \ \cite{BasoldGW17} via a type of signatures.
In this section, we give the definition of signatures, and then in the upcoming sections, we define the notion of HITs for a signature.

Higher inductive types are freely generated by both point and path constructors.
The data of the point constructor is a \emph{polynomial functor}.
These are described by the following type.

\begin{definition}
\label{def:poly}
The type $\mathcal{\poly}$ of \emph{codes of polynomials} is inductively generated by the following constructors.
\begin{center}
\begin{bprooftree}
\AxiomC{$X : \hset$}
\UnaryInfC{$\C \> X : \poly$}
\end{bprooftree}
\begin{bprooftree}
\AxiomC{$\I : \poly$}
\end{bprooftree}
\begin{bprooftree}
\AxiomC{$P : \poly$}
\AxiomC{$Q : \poly$}
\BinaryInfC{$\sumP{P}{Q} : \poly$}
\end{bprooftree}
\begin{bprooftree}
\AxiomC{$P : \poly$}
\AxiomC{$Q : \poly$}
\BinaryInfC{$\prodP{P}{Q} : \poly$}
\end{bprooftree}
\end{center}
\end{definition}

In the next section, we show that each code $P : \poly$ gives rise to a functor $\semP{P} : \functor{\hset}{\hset}$.
A higher inductive type $H$ should have a point constructor $c : \semP{P} \> H \rightarrow H$, which represents the introduction rule for points.

HITs also have an introduction rule for paths, and for those, we need to give the possible endpoints of paths.
Each path is given a universally quantified equation, which can possibly make use of the point constructor.
For this reason, the type of endpoint must depend on the data of the point constructor.
We also indicate the source and target of the equation, and both of them depend polynomially on the HIT being defined.
The type of endpoints is defined as an inductive type.

@@

The polynomial $A$ represents the point constructor and that explains the endpoint $\constr$.
The polynomials $P$ and $Q$ represent the source and target of the equation respectively.
If we have a set $X$ with a map $c : \semP{A} \> X \rightarrow X$, each endpoint $e : \ep{A}{P}{Q}$ gives rise to a natural map $\semP{P} \> X \rightarrow \semP{Q} \> X$ where $\constr$ is interpreted by $c$.
These maps are the left- and right-hand side of  the equations.

@@

\begin{definition}
\label{def:signature}
A \emph{HIT signature} $\sig$ consists of
\begin{itemize}
	\item A polynomial $\pointc{\sig} : \poly$ representing the point constructor;
	\item A type $\pathI{\sig}$ representing the names of path constructors;
	\item A family $\pathA{\sig} : \pathI{\sig} \rightarrow \poly$ representing the source of the paths;
	\item Maps $\pathlh{\sig}, \pathrh{\sig} : \deprod{j}{\pathA{\sig}}{\ep{\pointc{\sig}}{\pathA{\sig} \> j}{\I}}$ representing the left- and right-hand side of the equations.
\end{itemize}
\end{definition}

@@

\begin{lstlisting}[mathescape=true]
Inductive $H$ :=
| $c$ : $\semP{\pointc{\sig}} \> H \rightarrow H$
| $p$ : $\deprod{j}{\pathI{\sig}}{\deprod{x}{\semP{\pathA{\sig}} \> H}{\semE{\pathlh{\sig} \> j} \> H \> x = \semE{\pathrh{\sig} \> j} \> H \> x}}$
\end{lstlisting}

To illustrate the possibilities of this definition, we define two examples from the introduction as HIT signatures.
The first one is the integers modulo 2 and the second one is the propositional truncation of a set.

\begin{example}
\label{ex:modsig}
Recall $\modZ$ from the introduction.
We represent it by the signature $\modsig$ defined as follows
\begin{itemize}
	\item $\pointc{\modsig} \Def \sumP{\I}{(\C \> \unitt)}$;
	\item $\pathI{\modsig} \Def \unitt$;
	\item $\pathA{\modsig} \Def \I$;
	\item $\pathlh{\modsig} \> j \Def \comp{\comp{\comp{\comp{\id{\I}}{\inle}}{\constr}}{\inle}}{\constr}$;
	\item $\pathrh{\modsig} \> j \Def \id{\I}$;
\end{itemize}
\end{example}

Intuitively, this signature represents a HIT $H$ with operation $H + \unitt \rightarrow H$ representing the successor and zero.
Note that the endpoint $\comp{\inle}{\constr}$ takes the successor, so, the equation says that $S(S \> x) = x$ for all $x : X$.
Signatures for groups and join semi-lattices are defined similarly.
Since signatures can depend on types, we can also define the propositional truncation.

\begin{example}
\label{ex:truncsig}
Let $A$ be a set.
We represent the truncation of a set by the following signature.
\begin{itemize}
	\item $\pointc{\trunc} \Def \C \> A$;
	\item $\pathI{\trunc} \Def \unitt$;
	\item $\pathA{\trunc} \Def \prodP{\I}{\I}$;
	\item $\pathlh{\trunc} \> j \Def \comp{\id{\prodP{\I}{\I}}}{\prle}$;
	\item $\pathrh{\trunc} \> j \Def \comp{\id{\prodP{\I}{\I}}}{\prre}$;
\end{itemize}
\end{example}

This signature represents a HIT $H$ with a map $A \rightarrow H$ and for which all inhabitants are equal, which is precisely the propositional truncation from Definition \ref{def:trunc} for sets $A$.

@@

\subsection{Algebras in Sets}
Now we know what signatures are, the next goal is to define HITs for signatures.
For that we need to give the required introduction, elimination, and computation rules.
In this section, we study the introduction rule of HITs and we do that by defining a category of algebras on a polynomial.
Furthermore, since ultimately we want to construct HITs as quotients, we do not just define algebra in sets, but also in setoids.
Let us start by interpreting polynomials as functors.

>>>>

% Endpoints
\newcommand{\ep}[3]{\mathcal{E}_{#1}(#2,#3)} % type of endpoinst
\newcommand{\id}[1]{\constructor{id}_{#1}} % identity endpoints
\newcommand{\comp}[2]{#1 \cdot #2} % composition of endpoints
\newcommand{\inle}{\constructor{inl}} %left inclusion of endpoints
\newcommand{\inre}{\constructor{inr}} % right inclusion of endpoints
\newcommand{\prle}{\constructor{pr}_1} % first projection of endpoints
\newcommand{\prre}{\constructor{pr}_2} % second projection of endpoints
\newcommand{\pair}[2]{(#1 , #2)} % pairing of endpoints
\newcommand{\Ce}{\constructor{c}} % constant endpoint
\newcommand{\constr}{\constructor{constr}} % constructor endpoint
\newcommand{\fmap}{\constructor{fmap}} % function endpoint

@@

% Algebras
\newcommand{\semP}[1]{\llbracket #1 \rrbracket} % semantics of polynomials
\newcommand{\prealg}[1]{\category{PreAlg}_{\hset}(#1)} % category of prealgebras
\newcommand{\forget}[1]{\functortxt{U}_{#1}} % forgetful functor
\newcommand{\semE}[1]{\llbracket #1 \rrbracket} % semantics of endpoints

\newcommand{\Alg}[1]{\category{Alg}_{\hset}(#1)} % category of algebras
\newcommand{\constrA}[1]{\function{point}_{#1}} % point constructor of algebra
\newcommand{\constrP}[1]{\function{path}_{#1}} % path constructor of algebra

@@

% Examples of HITs
\newcommand{\Circ}{S^1} % circle
\newcommand{\base}{\constructor{base}} % base point of circle
\newcommand{\loopS}{\constructor{loop}} % base point of circle
\newcommand{\modZ}{\mathbb{Z}_2} % integers modulo 2
\newcommand{\ZZ}{\constructor{Z}} % zero
\newcommand{\SZ}{\constructor{S}} % successor
\newcommand{\modP}{\constructor{mod}} % n = n + 2
\newcommand{\setZ}{\constructor{Ztrunc}} % mod 2 is a set
\newcommand{\HIT}{\function{HIT}} % the HIT of a signature

@@

\def\lastname{Van der Weide, Geuvers}
\begin{document}
\begin{frontmatter}
  \title{The Construction of Set-Truncated Higher Inductive Types} \author{Niels van der Weide
		\thanksref{ALL}\thanksref{myemail}}
  \address{Institute for Computation and Information Sciences\\ Radboud Universiteit\\
  	Nijmegen, The Netherlands}
  \author{Herman Geuvers\thanksref{coemail}}
  \address{Institute for Computation and Information Sciences\\ Radboud Universiteit\\
  	Nijmegen, The Netherlands}
  \thanks[ALL]{The authors would like to thank Dan Frumin and Andrej Bauer for inspiring discussions. The authors also thank the reviewers for their helpful comments to improve the readability of this paper. Work on this article was supported by a grant from the \href{https://eutypes.cs.ru.nl/}{COST Action EUTypes CA15123}.} \thanks[myemail]{Email:
  	\href{mailto:nweide@cs.ru.nl} {\texttt{\normalshape
  			nweide@cs.ru.nl}}} \thanks[coemail]{Email:
  	\href{mailto:herman@cs.ru.nl} {\texttt{\normalshape
  			herman@cs.ru.nl}}}
\begin{abstract} 
  We construct finitary set-truncated higher inductive types (HITs) from quotients and the propositional truncation.
  For that, we first define signatures as a modification of the schema by Basold \etal, and we show they give rise to univalent categories of algebras in both sets and setoids.
  To interpret HITs, we use the well-known method of initial algebra semantics.
  The desired algebra is obtained by lifting the quotient adjunction to the level of algebras and adapting Dybjer's and Moeneclaey's interpretation of HITs in setoids.
  From this construction, we conclude that the equality types of HITs are freely generated and that HITs are unique.
  The results are formalized in the UniMath library.
\end{abstract}
\begin{keyword}
higher inductive types, homotopy type theory, category theory, setoids, intuitionistic type theory, Coq
\end{keyword}
\end{frontmatter}
\section{Introduction}
\label{intro}
Homotopy type theory (HoTT) is a form of intensional type theory. 
It has semantics in the simplicial sets model \cite{simpset}, and types represent spaces, terms represent points, and equalities represent paths, and so on.
Furthermore, equality is proof relevant and we frequently talk about homotopies: paths between paths.

One of the main features of HoTT is higher inductive types (HITs) \cite{hottbook}.
These are types generated by constructors for their points, paths, homotopies, and so on.
HITs have been used in numerous applications among which are homotopical patch theory \cite{AngiuliMLH16}, synthetic homotopy theory \cite{LicataF14,LicataS13,hottbook}, defining type theory within type theory \cite{AltenkirchK16,DBLP:journals/lmcs/AltenkirchK17}, constructive finiteness \cite{FruminGGW18}, and the partiality monad \cite{AltenkirchDK17}.
To get a feeling for what HITs are, let us look at some examples.

@@

\begin{lstlisting}[mathescape=true]
Inductive $\Circ$ :=
| $\base$ : $\Circ$
| $\loopS$ : $\base = \base$
\end{lstlisting}
&
\begin{lstlisting}[mathescape=true]
Inductive $\trunct{A}$ :=
| $\truncel{\cdot}$ : $A \rightarrow \trunct{A}$
| $p$ : $\deprod{x, y}{\trunct{A}}{x = y}$
\end{lstlisting}
&
\begin{lstlisting}[mathescape=true]
Inductive $\modZ$ :=
| $\ZZ$ : $\modZ$
| $\SZ$ : $\modZ \rightarrow \modZ$
| $\modP$ : $\deprod{x}{\modZ}{\SZ(\SZ \> x) = x}$
| $\setZ$ : $\isaset{\modZ}$
\end{lstlisting}
\end{tabular}

The first one, $\Circ$, is the \emph{circle}.
It has a point $\base$ and a path $\loopS : \base = \base$.
Since neither the point nor path constructors uses arguments from $\Circ$, this HIT is \emph{non-recursive}.
The second, $\trunct{A}$, is the \emph{propositional truncation} of $A$.
This type is $A$ with all its points identified.
Note that the path constructor $p$ uses arguments from $\trunct{A}$, which means this HIT is \emph{recursive}.
The last one, $\modZ$, is the \emph{integers modulo 2} and note that both the point constructor $\SZ$ and path constructor $\modP$ are recursive.
A HIT is called \emph{finitary} if its point constructor is described by a finitary polynomial and it is called \emph{set-truncated} if it is a set.

@@

\subsection{Background and Related Work}
The HITs we study in this paper, are set-truncated and recursion is allowed for both the point and path constructors.
It is a variation of the scheme by Basold \etal \ \cite{BasoldGW17}, which does not allow constructors for higher paths.
Other schemata of higher inductive types have already been defined.
\emph{W-suspensions}, developed by Sojakova \cite{Sojakova15}, allow defining HITs without recursive constructors, but these are not necessarily truncated.
The scheme by Dybjer and Moenclaey is similar to the one by Basold \etal, as it allows recursion for both the points and paths, but in addition, it supports constructors for homotopies.
Since types are $\omega$-groupoids \cite{van2011types,lumsdaine2009weak}, a type-theoretical version of these also provide a semantic specification of HITs \cite{DBLP:conf/tlca/HirschowitzHT15}.
At the moment, the definition which encompasses all these options, are higher inductive-inductive types (HIITs) \cite{KaposiK18} where induction-induction and paths in arbitrary dimensions are permitted.
These are a generalization of quotient inductive-inductive types (QIITs) developed by Altenkirch \etal \ \cite{AltenkirchCDKF18}.
Furthermore, Cavallo and Harper extend computational cubical type theory with a scheme for indexed cubical inductive types \cite{AngiuliHW17,CavalloH19}.

In addition to the schemata, some meta theory also has been developed.
Here we internally provide initial algebra semantics for our scheme meaning that we show initial implies induction.
Awodey \etal \ showed a stronger result for inductive types in intensional type theory \cite{AwodeyGS12,dybjer1994inductive}, namely that initiality is equivalent to induction, and Sojakova proved this for W-suspensions \cite{Sojakova15}.
For QIITs, initial algebra semantics has also been given \cite{AltenkirchCDKF18}.
Beside initial algebra semantics, Cavallo and Harper shows that computational cubical type theory, with cubical inductive types, satisfies canonicity \cite{AngiuliHW17,CavalloH19}.

The main result of this paper is about constructing a class of higher inductive types from simple ones.
Both Kraus and Van Doorn show that the propositional truncation can be constructed from non-recursive HITs \cite{Kraus16,Doorn16} and Rijke shows that every truncation can be constructed via non-recursive HITs \cite{rijke2017join}.
Note that these all are about truncations instead of a more general scheme.
However, contrary to our result, they do not rely on the restriction to set-truncated types.
Kaposi \etal \ \cite{KaposiKA19} show that can finitary QIITs can be constructed from one specific QIIT, but their construction relies on UIP.
From our construction, we deduce that the path space of each HIT is freely generated.
Kraus and Von Raumer obtained a similar property for the coequalizer by providing a nicer induction principle for its path types \cite{KrausRaumer}.

Reducing the existence of HITs to the existence of a small number of them, gives an approach to defining the semantics of HITs.
Other approaches to this problem have also been considered.
Coquand \etal \ interpret various higher inductive types, such as spheres, the torus, the truncation, and the pushout, in cubical type theory \cite{BezemCH13,CohenCHM16,CoquandHM18}.
All of these examples are not set-truncated and thus not covered by our scheme.
On the other hand, Lumsdaine and Shulman also study the semantics of higher inductive types using a semantic scheme (cell monads with parameters) and they prove existence in sufficiently nice Quillen model categories \cite{lumsdaine2017semantics}.
Dybjer and Moeneclaey give an interpretation of their scheme in the groupoid model \cite{DybjerM18,HofmannS94}.

@@

All material in this paper are formalized over the UniMath library \cite{UniMath}.
The proof of the main theorem is 4060 lines of code and the additional examples are 3278 lines of code.
The formalization can be found on \url{https://github.com/nmvdw/SetHITs}.

@@

\begin{definition}
\label{def:setoid}
A \emph{setoid} is a set $X$ together with an equivalence relation on $X$.
A \emph{setoid morphism} between two setoids  is a map between the underlying sets which preserves the equivalence relation.
The \emph{category $\setoids$ of setoids} is the category with setoids and setoid morphisms as objects and morphisms.
\end{definition}

@@

Recall that, If $R$ is an equivalence relation on $X$, each $R \> x \> y$ has to be a proposition.
Note that in the usual category of setoids, one would take a quotient of the setoid morphisms.
However, we refrain to do so, because in our construction, we do not need any additional equality.
If $X$ is a set and $R$ is an equivalence relation on $X$, then we write $(X, R)$ for the setoid with this data.
If no confusion arises, we write $x \equiv y$ for $R \> x \> y$.

We finish this section by giving some operations of functors and natural transformations, which we need in Definition \ref{def:semendpoint}.
Functors and natural transformations are defined the usual way \cite{mac2013categories}, and we write $\functor{\mathcal{C}}{\mathcal{D}}$ and $\nattrans{F}{G}$ for the type of functors from $\mathcal{C}$ to $\mathcal{D}$ and transformations from $F$ to $G$ respectively.

\begin{notation}
\label{def:functor}
We have the following functors.
\begin{itemize}
	\item For each category $\mathcal{C}$, we have the identity $\idf{\mathcal{C}} : \functor{\mathcal{C}}{\mathcal{C}}$.
	\item For $F : \functor{\mathcal{C}_1}{\mathcal{C}_2}$ and $G : \functor{\mathcal{C}_2}{\mathcal{C}_3}$, we have a composition $\compf{F}{G} : \functor{\mathcal{C}_1}{\mathcal{C}_3}$;
	\item If $\mathcal{D}$ has binary products, then for $F, G : \functor{\mathcal{C}}{\mathcal{D}}$, we have a product $\prodf{F}{G} : \functor{\mathcal{C}}{\mathcal{D}}$;
	\item If $\mathcal{D}$ has binary sums, then for $F, G : \functor{\mathcal{C}}{\mathcal{D}}$, we have a sum $\sumf{F}{G} : \functor{\mathcal{C}}{\mathcal{D}}$;
	\item For each object $X : \mathcal{D}$, we have a constant functor $\Cf{X} : \functor{\mathcal{C}}{\mathcal{D}}$.
\end{itemize}
\end{notation}

@@

The type of $\pairt{\eta_1}{\eta_2}$ might seem unnatural.
However, it is precisely this type we need for the constructions in Definition \ref{def:semendpoint}..

\section{Signature of HITs}
\label{sec:signature}
Before we study the construction of set truncated HITs, we first give a precise definition of those.
We describe HITs by saying how to construct points and paths.
For example, the integers modulo 2, which we considered in the introduction, has two constructors for points, namely $\ZZ$ ad $\SZ$, and one for the paths, namely $\modP$.
The signature of this type would thus indicate that we have one nullary and one unary operation, and a path of the required type.
In this section, we define a type of signatures as internalized variation of the one by Basold \etal \ \cite{BasoldGW17}, and then in the upcoming sections, we define HITs for a signature.

The first ingredient of the signature, is the arity of the point constructor.
This is described by a \emph{finitary polynomial functor}, and we define those as the following type.

\begin{definition}
\label{def:poly}
We define the type $\mathcal{\poly}$ of \emph{codes of finitary polynomials} as the inductive type generated by the following constructors.
\begin{center}
\begin{bprooftree}
\AxiomC{$X : \hset$}
\UnaryInfC{$\C \> X : \poly$}
\end{bprooftree}
\begin{bprooftree}
\AxiomC{$\I : \poly$}
\end{bprooftree}
\begin{bprooftree}
\AxiomC{$P : \poly$}
\AxiomC{$Q : \poly$}
\BinaryInfC{$\sumP{P}{Q} : \poly$}
\end{bprooftree}
\begin{bprooftree}
\AxiomC{$P : \poly$}
\AxiomC{$Q : \poly$}
\BinaryInfC{$\prodP{P}{Q} : \poly$}
\end{bprooftree}
\end{center}
\end{definition}

In the next section, we show that each code $P : \poly$ gives rise to a functor $\semP{P} : \functor{\hset}{\hset}$.
Each higher inductive type $H$ has a point constructor $c : \semP{P} \> H \rightarrow H$, which represents the introduction rule for points.

HITs also have an introduction rule for paths, and for those, we need to give the possible endpoints of paths.
Each path is given by a universally quantified equation, which can possibly make use of the point constructor.
For this reason, the type of endpoint must depend on the data of the point constructor.
We also indicate the source and target of the equation, and both of them depend polynomially on the HIT being defined.
The type of endpoints is defined inductively.

@@

In the formalization, we also include for each $f : X \rightarrow Y$ an endpoint $\fmap \> f : \ep{A}{\C \> X}{\C \> Y}$, which we do not discuss here.
The polynomial $A$ represents the point constructor and that explains the endpoint $\constr$.
The polynomials $P$ and $Q$ represent the source and target of the equation respectively.
If we have a set $X$ with a map $c : \semP{A} \> X \rightarrow X$, each endpoint $e : \ep{A}{P}{Q}$ gives rise to a natural morphism from the functor $\semP{P}$ to $\semP{Q}$ where $\constr$ is interpreted using $c$.
These maps are the left- and right-hand side of  the equations.

@@

\begin{definition}
\label{def:signature}
A \emph{HIT signature} $\sig$ consists of
\begin{itemize}
	\item A polynomial $\pointc{\sig} : \poly$ representing the point constructor;
	\item A type $\pathI{\sig}$ representing the names of path constructors;
	\item A family $\pathA{\sig} : \pathI{\sig} \rightarrow \poly$ representing the arguments of the paths;
	\item Maps $\pathlh{\sig}, \pathrh{\sig} : \deprod{j}{\pathI{\sig}}{\ep{\pointc{\sig}}{\pathA{\sig} \> j}{\I}}$ representing the left- and right-hand side of the equations.
\end{itemize}
\end{definition}

@@

\begin{lstlisting}[mathescape=true]
Inductive $H$ :=
| $\constructor{c}$ : $\semP{\pointc{\sig}} \> H \rightarrow H$
| $\constructor{p}$ : $\deprod{j}{\pathI{\sig}}{\deprod{x}{\semP{\pathA{\sig} \> j} \> H}{\semE{\pathlh{\sig} \> j} \> H \> x = \semE{\pathrh{\sig} \> j} \> H \> x}}$
| $\constructor{s}$ : $\isaset{H}$
\end{lstlisting}

Since our goal is to interpret set-truncated HITs, we require the constructor $s$.
To illustrate the possibilities of this definition, we define two examples from the introduction as HIT signatures.
The first one is the integers modulo 2 and the second one is the propositional truncation of a set.

\begin{example}
\label{ex:modsig}
Recall $\modZ$ from the introduction.
We represent it by the signature $\modsig$ defined as follows
\[
\pointc{\modsig} \Def \sumP{\I}{(\C \> \unitt)}
\quad \quad \quad
\pathI{\modsig} \Def \unitt
\quad \quad \quad
\pathA{\modsig} \Def \I\]
\[
\pathlh{\modsig} \> j \Def \comp{\comp{\comp{\inle}{\constr}}{\inle}}{\constr}
\quad \quad \quad
\pathrh{\modsig} \> j \Def \id{\I}
\]
\remove{
\begin{itemize}
	\item $\pointc{\modsig} \Def \sumP{\I}{(\C \> \unitt)}$;
	\item $\pathI{\modsig} \Def \unitt$;
	\item $\pathA{\modsig} \Def \I$;
	\item $\pathlh{\modsig} \> j \Def \comp{\comp{\comp{\inle}{\constr}}{\inle}}{\constr}$;
	\item $\pathrh{\modsig} \> j \Def \id{\I}$;
\end{itemize}
}
\end{example}

Intuitively, this signature represents a HIT $H$ with operation $H + \unitt \rightarrow H$ representing the successor and zero.
Note that the endpoint $\comp{\inle}{\constr}$ takes the successor, so, the equation says that $S(S \> x) = x$ for all $x : X$.
Since signatures can depend on types, we can also define the propositional truncation.

\begin{example}
\label{ex:truncsig}
Let $A$ be a set.
We represent its truncation by the following signature.
\[
\pointc{\trunc} \Def \C \> A
\quad \quad \quad
\pathI{\trunc} \Def \unitt
\quad \quad \quad
\pathA{\trunc} \Def \prodP{\I}{\I}
\]
\[
\pathlh{\trunc} \> j \Def \prle
\quad \quad \quad
\pathrh{\trunc} \> j \Def \prre
\]
\remove{
\begin{itemize}
	\item $\pointc{\trunc} \Def \C \> A$;
	\item $\pathI{\trunc} \Def \unitt$;
	\item $\pathA{\trunc} \Def \prodP{\I}{\I}$;
	\item $\pathlh{\trunc} \> j \Def \prle$;
	\item $\pathrh{\trunc} \> j \Def \prre$;
\end{itemize}
}
\end{example}

Other examples, such as the integers, free algebras, and the ring of polynomials, can be found in the formalization.
This signature represents a HIT $H$ with a map $A \rightarrow H$ and for which all inhabitants are equal, which is precisely the propositional truncation from Definition \ref{def:trunc} for sets $A$.

@@

\subsection{Algebras in Sets}
To define HITs on a signature, we need to give the introduction, elimination, and compuation rules.
Let us start with the introduction rule, which we define via algebras.
Note that this rule comes in two flavors: one for the points and one forthe paths.
For this reason, we define algebras in two steps.

We start by saying how to interpret the point constructors.
